Causal Inference in Population Trends: Searching for Demographic Anomalies in Big Data

### 2021-03-29 MainDocument/MainDocument.Rmd
Updated.

authors: 
- name: Mathew E. Hauer
  thanks:  Thanks y'all!
  affiliation: Department of Sociology, Florida State University

@@

abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in social data, previously dismissed as noise. Here, we combine causal inference techniques and abductive reasoning to identify fertility and mortality anomalies on twenty years of complete demographic data in the United States. We uncover real, “hidden” baby booms/busts and mortality spikes/dips, distinguishable from regular trend variations. We identify more than 22 and 156 fertility and mortality anomalies, totaling more than 200k and 600k anomalous births and deaths, respectively. Notable detectable mortality anomalies include the September 11 2001 terrorist attack in New York, the emergence and acceleration of the opioid epidemic in New Hampshire, and the excess mortality due to Hurricane Maria in Puerto Rico. Notable fertility anomalies include the “missing births” in Louisiana after Hurricane Katrina and the reduction in fertility behavior after the September 2008 stock market crash in Connecticut, amongst others. Our combined causal inference and abductive reasoning approach can be readily adapted to find other, undiscovered social phenomena or to evaluate the efficacy of important public policies."

@@

@@

The search for causality is one of the primary motivations of social scientists [@smith2011commentary; @pearl2009causal]. Causal inference, the identification of causal mechanisms, attempts to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp]. Often, the identification of a casual mechanism is accomplished via a randomized control trial (RCT) or expert knowledge applied to a natural experiment [@salganik2019bit].

@@

Additionally, both RCTs and natural experiments are based on traditional hypothesis-based testing, the deductive scientific paradigm where a question is first posed and scientists generate or find data to answer the question. Such approaches, while extensively utilized and time-tested, increase the likelihood that a researcher will miss important phenomena that might go unnoticed because it falls outside the hypothesis, or the study could suffer from extensive p-hacking [@head2015extent;@nuzzo2014scientific;@ruggles2014big]. At the same time, purely inductive data mining can result in false assertions or spurious findings, especially in the hands of those who are not field experts [@yanai_hypothesis_2020]. Abductive reasoning, sometimes described as “inferring cause from effect” [@Crowder2017], moves the scientific process from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] Physical scientists use abductive reasoning more commonly than social scientists. The discovery of Neptune in 1846 based on unexplained perturbations of the orbit of Uranus [@popper2005logic], the discovery of galactic migration of planets [@gomes2005n], and the current generation of gravitational wave detectors such as the Laser Interferometer Gravitational-Wave Observatory (LIGO) [@harry2010advanced] demonstrate the power of the abductive approach in the physical sciences.

@@

In this paper, we combine a modern statistical outlier detection algorithm [@chen1993joint], abductive reasoning, and large demographic data sets on more than 20 years of mortality and fertility trends to identify anomalous demographic behavior. In essence, we identify *effects without knowledge of the cause*. We ask one fundamental question regarding demographic anomalies: What are the hidden baby booms/busts and mortality spikes/dips in the United States over the last twenty years? We do not necessarily know the causes of these anomalies but identifying them allows scientists with more detailed knowledge of local population dynamics, state-level policy making, or macro-economics to explain these phenomena post-hoc. From explanations of phenomena that may have previously gone unnoticed, social scientists may be able to better forecast populations and provide policy solutions for impending problems such as climate change.

@@

## Robustness Check
To determine the false-positive (Type I) and false-negative (Type II) error rates under a critical value of 3.5, we ran two simulations. For the false-positive error we simulated 1000 time series with 180 time steps without introducing any anomalies and ran the `tsoutliers` algorithm on them. Any anomalies we detect in these time series are spurious anomalies. For the false-negative error, we simulated 1000 time series with 180 time steps but introduced a single anomaly of the same magnitude at the same time point in each time series. If the `tsoutliers` algorithm fails to detect the anomaly, it would be considered a false-negative error.

```{r , include = FALSE, cache = TRUE}

set.seed(1)
results_noutliers <- data.frame()
# y2 <- matrix(nrow = 1000, ncol=180)
for(i in 1:1000){
  tryCatch({
y <- arima.sim(model = list(ar = 0.7, ma = -0.4), n = 180)

df2<- tso(y, cval = 3.5)$outliers
df2$iteration <- i
results_noutliers <- bind_rows(results_noutliers, df2)
# print(i)
  } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

# (a <- nrow(results_noutliers) / (180*1000))
# 
# a * 20000

results_AO <- data.frame()
# y2 <- matrix(nrow = 1000, ncol=180)
for(i in 1:1000){
  tryCatch({
    y <- arima.sim(model = list(ar = 0.7, ma = -0.4), n = 180)
    y[150] <- 4

    df2<- tso(y, cval = 3.5)$outliers
    df2$iteration <- i
    results_AO <- bind_rows(results_AO, df2)
    # print(i)
  } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}
```
In our sans-anomaly time series, we detect `r nrow(results_noutliers)` anomalies on the 180,000 candidate time steps for a false-positive error rate of `r prettyNum(nrow(results_noutliers) / 180000*100, digits = 3)`%. In our time series containing a single anomaly, we detect `r nrow(results_AO[which(results_AO$time == 150),])` anomalies out of the 1000 candidate time series, for a false-negative error rate of `r (1-(nrow(results_AO[which(results_AO$time == 150),]) / 1000)) * 100`%. This provides a statistical power of `r nrow(results_AO[which(results_AO$time == 150),]) / 1000`. These results give us confidence that the anomalies we detect in this paper are true anomalies.

![**"Toy" example of anomaly detection using the exchange rate between the Euro and the Swiss Franc.** The Swiss Franc was pegged to the Euro at 1.2 until January 15 2015 when the peg was removed and the currency was allowed to trade more freely. This is a very strong anomly in the time series (*t*-stat = -144.07) detectable using statistical methods and visually. The countefactural estimate with the outlier removed would simply have kept the currency exchange at 1.2 ($\hat{y}$). \label{explanationfigure}](../MainDocument/MainDocument_files/figure-latex/Figure1.png)

@@

stockdat <- tribble(
  ~Date, ~Value,
'2014-12-27',	1.203880343,
'2014-12-28',	1.202523908,
'2014-12-29',	1.202714098,
'2014-12-30',	1.201988512,
'2014-12-31',	1.202535102,
'2015-01-01',	1.201727726,
'2015-01-02',	1.201794127,
'2015-01-04',	1.20132752,
'2015-01-05',	1.20100032,
'2015-01-06',	1.20071649,
'2015-01-07',	1.200762136,
'2015-01-08',	1.200787933,
'2015-01-09',	1.20079581,
'2015-01-11',	1.200989867,
'2015-01-12',	1.2006267,
'2015-01-13',	1.20058359,
'2015-01-14',	1.200605219,
'2015-01-15',	0.99417984,
'2015-01-16',	0.992258896,
'2015-01-18',	0.992353824,
'2015-01-19',	1.019880601,
'2015-01-20',	1.00940244,
'2015-01-21',	0.999117945,
'2015-01-22',	0.988725725,
'2015-01-23',	0.939377117,
'2015-01-24',	0.987456601,
'2015-01-25',	0.980699544,
'2015-01-26',	1.013809279,
'2015-01-27',	1.02203072,
'2015-01-28',	1.0234588,
'2015-01-29',	1.0455615
)
dat3 <- ts(stockdat$Value)
stockdat_outlier <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)
```

@@

## Toy example
**\autoref{explanationfigure}** shows a toy example for anomaly detection in a time series using the classic example of the exchange rate between the Euro and the Swiss Franc (CHF). On January 15 2015, the Swiss National Bank removed the currency peg of 1.20 francs per Euro, exposing the Franc to the volatility of the currency market. The Franc immediately began trading a reduced rate compared to the Euro. What would have been the Euro/CHF exchange rate had the Swiss National Bank *not* removed the peg? A simple counter-factual estimate would be to keep the exchange rate at 1.20 ($\hat{y}$, sometimes called a "synthetic control" [@abadie2010synthetic]).

In the Swiss Franc example, we have knowledge of the Swiss Bank's activities after the fact or ex-post-facto to to create the counter-factual time series $\hat{y}$ of 1.20. But is this anomaly detectable without knowledge of the Swiss Bank's activities? In other words, can we detect the reduction of the exchange rate on January 15 using only the time series? Absolutely. the tsoutliers package identifies January 15 2015 as an extremely strong level-shift outlier (*t*-stat = `r round( stockdat_outlier$outliers$tstat[1], digits = 2)`). The real world is hardly this simplified where a direct intervention is known and is testable.

@@

## Replication Files
All data and code necessary to reproduce the reported results are licensed under the CC-BY-4.0 license and are publicly available in a replication repository located at https://osf.io/fp52x/?view_only=754d9a72a2ea4f6b8e0c193dc9a590d1. The analyses were performed in *R* [@rcite].

# Results
We detect numerous anomalous mortality and fertility events at the US state-level since 1999. A full listing of these anomalies can be found in the **Supplementary Materials**. We begin by highlighting significant fertility and mortality anomalies across all three types of outliers (Additive Outliers, Level Shift Outliers, and Temporary Change outliers) with plausible explanations. We then highlight two strong anomalies that bely explanation. Finally, we conclude with a summary of the anomalies we detect.

@@

conn_tstat <- prettyNum(outlier.fert.conn$outliers$tstat, digits = 3)
conn_births <- prettyNum(outlier.fert.conn$outliers$coefhat, digits = 0)
conn_tot <- prettyNum(abs(sum(outlier.fert.conn$y - outlier.fert.conn$yadj)), digits = 0, big.mark = ",")
conn_per <- percent(abs(sum(outlier.fert.conn$y - outlier.fert.conn$yadj) / sum(outlier.fert.conn$yadj[80:180])))
```

@@

```{r FertilityAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.} The top part of each panel contains the original time series (light gray), the corrected, counter-factual time series in the absence of anomalies (blue), and the red dots correspond to the onset of detected anomalies. The bottom part of each panel contains the magnitude and type of the outlier in red. In (a), we detect two outliers, back-to-back, likely resulting from Hurricane Katrina in August and October 2005, representing a decrease of more than 4,400 births due to the hurricane (t-statistics= -6.002 and -4.997). In (b), we detect one outlier, a level shift outlier (LS) in August 2009. We believe this reduction is attributable to the stock market crash 11 months earlier (t-statistic = -3.71). \\label{fig:fertla}")}

a <-  function() {
  plot(outlier.fert.la, cex.axis=0.8)
  title(main = "Louisiana - Fertility", adj = 0.9)
}

b <- function(){
  plot(outlier.fert.conn, cex.axis=0.8) +
  title(main = "Connecticut - Fertility", adj = 0.88)
}

plot_grid(ggdraw(a) ,
               ggdraw(b), 
               ncol=1,
               labels= "auto")
```

>>>>

authors: 
- name: Mathew E. Hauer
  # thanks:  Thanks y'all!
  affiliation: Department of Sociology, Florida State University

@@

abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in social data, previously dismissed as noise. Here, we combine causal inference techniques and abductive reasoning to identify fertility and mortality anomalies on twenty years of complete demographic data in the United States. We uncover real, “hidden” baby booms/busts and mortality spikes/dips, distinguishable from regular trend variations. We identify more than 22 and 156 fertility and mortality anomalies, totaling more than 200k and 600k anomalous births and deaths, respectively. Notable detectable mortality anomalies include the September 11 2001 terrorist attack in New York and the emergence and acceleration of the opioid epidemic in New Hampshire. Notable fertility anomalies include the “missing births” in Louisiana after Hurricane Katrina and the reduction in fertility behavior after the September 2008 stock market crash in Connecticut, amongst others. The combined causal inference and abductive reasoning approach can be readily adapted to find other, undiscovered social phenomena or to evaluate the efficacy of important public policies."

@@

 countylist = unique(dat_fert$location)

noamom_fert <- setdiff(countylist, unique(anom_fert$location))

noamom_mort <- setdiff(countylist, unique(anom_mort$location))

  results_noutliers <- read_csv("../R/DATA-PROCESSED/robustness_nooutliers.csv")
  results_AO <-   read_csv("../R/DATA-PROCESSED/robustness_AO.csv")

z2 <- results_noutliers %>%
  group_by(type) %>%
  dplyr::summarise(tot = n()) %>%
  mutate(timesteps = if_else(type == "fertility", 180*1000*33, 228*1000*8)) %>%
  dplyr::summarise(tot = sum(tot),
                   timesteps = sum(timesteps)) %>%
  mutate(FNR = tot/timesteps)

@@

The search for causality is one of the primary motivations of social scientists [@smith2011commentary; @pearl2009causal]. Causal inference, the identification of causal effects, attempts to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp]. Often, the identification of a casual mechanism is accomplished via a randomized control trial (RCT) or expert knowledge applied to a natural experiment [@salganik2019bit].

@@

Additionally, both RCTs and natural experiments are based on traditional hypothesis-based testing, the deductive scientific paradigm where a question is first posed and scientists generate or find data to answer the question. Such approaches, while extensively utilized and time-tested, increase the likelihood that a researcher will miss important phenomena that might go unnoticed because it falls outside the hypothesis, or the study could suffer from extensive p-hacking [@head2015extent;@nuzzo2014scientific;@ruggles2014big]. At the same time, purely inductive data mining can result in false assertions or spurious findings, especially in the hands of those who are not field experts [@yanai_hypothesis_2020]. Abductive reasoning, sometimes described as “inferring cause from effect” [@Crowder2017], moves the scientific process from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm]. Physical scientists use abductive reasoning more commonly than social scientists. The discovery of Neptune in 1846 based on unexplained perturbations of the orbit of Uranus [@popper2005logic], the discovery of galactic migration of planets [@gomes2005n], and the current generation of gravitational wave detectors such as the Laser Interferometer Gravitational-Wave Observatory (LIGO) [@harry2010advanced] demonstrate the power of the abductive approach in the physical sciences.

@@

In this paper, we combine a modern statistical outlier detection algorithm [@chen1993joint], abductive reasoning, and large demographic data sets on more than 20 years of mortality and fertility trends to identify anomalous demographic behavior. In essence, we identify *effects without knowledge of the cause*. We ask one fundamental question regarding demographic anomalies: What are the hidden baby booms/busts and mortality spikes/dips in the United States over the last twenty years? We do not necessarily know the causes of these anomalies and suggest causes for select anomalies. Even without knowledge of the causes, simply identification of anomalies allows scientists with more detailed knowledge of local population dynamics, state-level policy making, or macro-economics to explain these phenomena post-hoc. From explanations of phenomena that may have previously gone unnoticed, social scientists may be able to better forecast populations and provide policy solutions for impending problems such as climate change.

@@

<!-- To determine the false-positive (Type I) and false-negative (Type II) error rates under a critical value of 3.5, we ran two simulations. For the false-positive error we simulated 1000 time series with 180 time steps without introducing any anomalies and ran the `tsoutliers` algorithm on them. Any anomalies we detect in these time series are spurious anomalies. For the false-negative error, we simulated 1000 time series with 180 time steps but introduced a single anomaly of the same magnitude at the same time point in each time series. If the `tsoutliers` algorithm fails to detect the anomaly, it would be considered a false-negative error. -->

<!-- ```{r , include = FALSE, cache = TRUE} -->

<!-- set.seed(1) -->
<!-- results_noutliers <- data.frame() -->
<!-- # y2 <- matrix(nrow = 1000, ncol=180) -->
<!-- for(i in 1:1000){ -->
<!--   tryCatch({ -->
<!-- y <- arima.sim(model = list(ar = 0.7, ma = -0.4), n = 180) -->

<!-- df2<- tso(y, cval = 3.5)$outliers -->
<!-- df2$iteration <- i -->
<!-- results_noutliers <- bind_rows(results_noutliers, df2) -->
<!-- # print(i) -->
<!--   } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")}) -->
<!-- } -->

<!-- # (a <- nrow(results_noutliers) / (180*1000)) -->
<!-- #  -->
<!-- # a * 20000 -->

<!-- results_AO <- data.frame() -->
<!-- # y2 <- matrix(nrow = 1000, ncol=180) -->
<!-- for(i in 1:1000){ -->
<!--   tryCatch({ -->
<!--     y <- arima.sim(model = list(ar = 0.7, ma = -0.4), n = 180) -->
<!--     y[150] <- 4 -->

<!--     df2<- tso(y, cval = 3.5)$outliers -->
<!--     df2$iteration <- i -->
<!--     results_AO <- bind_rows(results_AO, df2) -->
<!--     # print(i) -->
<!--   } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")}) -->
<!-- } -->
<!-- ``` -->
<!-- In our sans-anomaly time series, we detect `r nrow(results_noutliers)` anomalies on the 180,000 candidate time steps for a false-positive error rate of `r prettyNum(nrow(results_noutliers) / 180000*100, digits = 3)`%. In our time series containing a single anomaly, we detect `r nrow(results_AO[which(results_AO$time == 150),])` anomalies out of the 1000 candidate time series, for a false-negative error rate of `r (1-(nrow(results_AO[which(results_AO$time == 150),]) / 1000)) * 100`%. This provides a statistical power of `r nrow(results_AO[which(results_AO$time == 150),]) / 1000`. These results give us confidence that the anomalies we detect in this paper are true anomalies. -->

<!-- ![**"Toy" example of anomaly detection using the exchange rate between the Euro and the Swiss Franc.** The Swiss Franc was pegged to the Euro at 1.2 until January 15 2015 when the peg was removed and the currency was allowed to trade more freely. This is a very strong anomly in the time series (*t*-stat = -144.07) detectable using statistical methods and visually. The countefactural estimate with the outlier removed would simply have kept the currency exchange at 1.2 ($\hat{y}$). \label{explanationfigure}](../MainDocument/MainDocument_files/figure-latex/Figure1.pdf){ width=100% height=40%} -->

@@

# stockdat <- tribble(
#   ~Date, ~Value,
# '2014-12-27',	1.203880343,
# '2014-12-28',	1.202523908,
# '2014-12-29',	1.202714098,
# '2014-12-30',	1.201988512,
# '2014-12-31',	1.202535102,
# '2015-01-01',	1.201727726,
# '2015-01-02',	1.201794127,
# '2015-01-04',	1.20132752,
# '2015-01-05',	1.20100032,
# '2015-01-06',	1.20071649,
# '2015-01-07',	1.200762136,
# '2015-01-08',	1.200787933,
# '2015-01-09',	1.20079581,
# '2015-01-11',	1.200989867,
# '2015-01-12',	1.2006267,
# '2015-01-13',	1.20058359,
# '2015-01-14',	1.200605219,
# '2015-01-15',	0.99417984,
# '2015-01-16',	0.992258896,
# '2015-01-18',	0.992353824,
# '2015-01-19',	1.019880601,
# '2015-01-20',	1.00940244,
# '2015-01-21',	0.999117945,
# '2015-01-22',	0.988725725,
# '2015-01-23',	0.939377117,
# '2015-01-24',	0.987456601,
# '2015-01-25',	0.980699544,
# '2015-01-26',	1.013809279,
# '2015-01-27',	1.02203072,
# '2015-01-28',	1.0234588,
# '2015-01-29',	1.0455615
# )
# dat3 <- ts(stockdat$Value)
# stockdat_outlier <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)

# download.file("https://seer.cancer.gov/popdata/yr1969_2019.19ages/la.1969_2019.19ages.txt.gz", "./R/DATA-RAW/la.1969_2019.19ages.txt.gz")
# ## UNZIPPING THE DATA FILE
# gunzip("./R/DATA-RAW/la.1969_2019.19ages.txt.gz", overwrite = TRUE, remove = TRUE)

K05_pop<- read.table("../R/DATA-RAW/la.1969_2019.19ages.txt") 
K05_pop$V1 <- as.character(K05_pop$V1) # SETTING THE ENTIRE SINGLE VARIABLE INTO A CHARACTER
K05_pop$YEAR <- as.numeric(substr(K05_pop$V1,1,4)) # SEPARATING THE YEAR AND SETTING IT AS A NUMBER
K05_pop$STATEID <- substr(K05_pop$V1, 5,6) # SEPARATING THE 2 CHARACTER STATE ABBREVIATION
K05_pop$STATE <- substr(K05_pop$V1, 7,8) # SEPARATING THE 2-DIGIT STATE CODE
K05_pop$COUNTY <- substr(K05_pop$V1,9,11) # SEPARATING THE 3-DIGIT COUNTY CODE
K05_pop$REGISTRY <- substr(K05_pop$V1, 12,12) # REGISTRY IS A THROW AWAY VARIABLE REFERING TO ODD GEOGRAPHIES
K05_pop$RACE <- substr(K05_pop$V1, 14,14) # SEPARATING OUT THE RACE CODES.
K05_pop$ORIGIN <- substr(K05_pop$V1, 15,15) # SEPARATING OUT HISPANIC ORIGIN. THIS VARIABLE IS NOT APPLICABLE IN THE 1969-2016 DATA
K05_pop$SEX <- substr(K05_pop$V1, 16,16) # SEPARATING OUT THE SEX DATA
K05_pop$POPULATION <- as.numeric(substr(K05_pop$V1, 19, 30)) # SEPARATING THE ACTUAL POPULATION ESTIMATES.

# Setting the groupings
GROUPING <- c("STATE", "COUNTY", "YEAR")

K05_pop <- K05_pop %>%
  group_by(.dots = GROUPING) %>%
  dplyr::summarise(POPULATION = sum(POPULATION))

start <- 1980
K05_pop$GEOID <- paste0(K05_pop$STATE, K05_pop$COUNTY) # SETTING THE 5-DIGIT FIPS CODE
K05_pop2 <- filter(K05_pop, GEOID == "22071",
                  YEAR >= start)

dat3 <- ts(K05_pop2$POPULATION)
a<- tsoutliers::tso(dat3,types = c("AO","LS","TC"))

K05_pop2$label = ifelse(K05_pop2$YEAR == 2006, "Anomaly Detected\n t = -96.4", "")

b <- as.data.frame(forecast(auto.arima(K05_pop2$POPULATION[1:26]), h=14)) %>%
  mutate(YEAR = seq(2006, 2019,1)) %>%
  dplyr::select(Point.Forecast = `Point Forecast`, YEAR)

b<- rbind(b, data.frame(YEAR = 2005, `Point Forecast` = 494294))

```{r ToyExample, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{An example of anomaly detection using Hurricanes Katrina and Rita in Orleans Parish Louisiana in 2005.} Hurricane Katrina struck Louisiana in 2005 and we use it as a toy example to illustrate our approach. This figure shows the annual time series of total population in Orleans Parish between 1980 and 2019. Between 1990 and 2005, Orleans Parish total population changed from 495k to 494k, suggesting a possible 'plateau' in the population (illustrated with the dotted 'counterfactual'). Hurricane Katrina and the widespread population loss of more than 200k people represent a very strong anomaly (t=-96.39) \\label{explanationfigure}")}

ggplot(data = K05_pop2, aes( x = YEAR, y = POPULATION)) +
  geom_line() +
  geom_point(data = K05_pop2[which(K05_pop2$label == "Anomaly Detected\n t = -96.4"),], aes(size=2), shape=1) + 
  geom_line(data = b, aes(y = Point.Forecast, x=YEAR), linetype=2) +
  scale_y_continuous(label=comma) +
  geom_label_repel(aes(label = label),
                   box.padding   = 0.4, 
                   point.padding = 0.6,
                   segment.color = 'grey50') +
  annotate(geom='text', x=2021, y=500000, label=TeX("Counterfactual estimate, $\\hat{y}$", output='character'), parse=TRUE, hjust = 0) +
  annotate(geom='text', x=2021, y=390000, label=TeX("Observed data, $y$", output='character'), parse=TRUE, hjust = 0) +
  scale_x_continuous(breaks = seq(1980, 2020, 10), limits = c(1980,2040), expand= c(0,0)) +
   scale_y_continuous(label=comma, limits=c(0, 565000), breaks = seq(0,570000,100000)) +
  coord_capped_cart(bottom='right') +
  theme_classic() +
  theme(legend.position = "none")+
  labs(x = "Year",
       y = "Total Population")

@@

## Toy example
**\autoref{explanationfigure}** shows a toy example for anomaly detection in a time series using the example of Hurricane Katrina in Orleans Parish. On August 23 2005, Hurricane Katrina, a category 5 hurricane, struck southern Louisiana causing widespread damage and destruction in Orleans Parish in particular. The displacement from the Hurricane and the federal response were well documented [@horiDisplacementDynamicsSouthern2009; @fussellRecoveryMigrationCity2014] and Census estimates suggest Orleans Parish lost more than 200,000 residents between 2005 and 2006. What would have been Orleans' population estimate had Hurricane Katrina *not* occurred? A simple counter-factual estimate might keep population size just under 500,000 people ($\hat{y}$).

In the Hurricane Katrina example, we have knowledge of the impact of Hurricane Katrina after the fact or ex-post-facto in order to create the counter-factual time series $\hat{y}$. But is this anomaly detectable without knowledge of the Hurricane Katrina? In other words, can we detect the reduction of Orleans Parish's population between 2005 and 2006 using only the time series? Absolutely. the tsoutliers package identifies 2006 as an extremely strong level-shift outlier (*t*-stat = `r round( a$outliers$tstat[1], digits = 2)`). The real world is hardly this simplified where a direct intervention is known, is large, and is testable.

@@

## Robustness Check
Chen and Liu [-@chen1993joint] report results of a simulation exercise of varying critical value levels. They simulated various anomaly-free time series and injected specific anomalies of varying $\sigma$ levels.  For a critical value of 3.5 -- the level Chen and Liu recommend to minimize both false-positive (Type I) and false-negative (Type II) error rates -- they report Type I error rates between 0.0\% and 0.1\% and Type II error rates or the statistical power between 40\% and 60\% with injected anomalies equal to 4$\sigma$. We follow their lead by simulating time series to test for Type I and Type II errors. We base our simulation on the `r length(noamom_fert) + length(noamom_mort)` time series that do not contain any anomalies (see **Results** below; `r length(noamom_fert)` without a single fertility anomaly and `r length(noamom_mort)` without a single mortality anomaly).

To test for false-positive rates, we simulate 1000 time series for each of the `r length(noamom_fert) + length(noamom_mort)` state-components (fertility or mortality) using the fitted ARIMA model for each state-component. This allows the simulated time series to mimic the underlying true time series (see **Supplementary Material** for the underlying code). Any anomalies we detect in the `r prettyNum(z2$timesteps, big.mark=',')` candidate time steps are spurious anomalies. We detect `r prettyNum(z2$tot, big.mark=',')` anomalies for a false-positive error rate of `r percent(z2$FNR, accuracy = 0.001)`.

To test for false-negative rates, we use the underlying true, sans-anomaly time series but introduce a single anomaly (magnitude equal to 3.5$\sigma$) at the same time point in each series. If the `tsoutliers` algorithm fails to detect the anomaly, it would be considered a false-negative error. We detect `r  prettyNum(nrow(results_AO[which(results_AO$time == 150),]), big.mark=',')` out of `r prettyNum((length(noamom_fert) + length(noamom_mort))*1000,, big.mark=',')` candidate time series giving us a false-negative error rate of `r percent((1-(nrow(results_AO[which(results_AO$time == 150),]) / ((length(noamom_fert) + length(noamom_mort))*1000))), 0.1)`. This provides a statistical power of `r percent(((nrow(results_AO[which(results_AO$time == 150),]) / ((length(noamom_fert) + length(noamom_mort))*1000))), 0.1)`.

Based on these tests we think it is clear that our probability of committing a Type I error (or claiming an anomaly is true when in fact is not) is very low. But, based on our false-negative error rate, it is possible that the anomalies we do detect are *conservative*. Our statistical power is of just `r percent(((nrow(results_AO[which(results_AO$time == 150),]) / ((length(noamom_fert) + length(noamom_mort))*1000))), 0.1)` suggests a higher than normal likelihood that true effects are going undetected. We are unlikely to commit an error of commission (ie, including a false anomaly) but are more likely to commit an error of omission (ie omitting true anomalies). Thus our results should be considered conservative.

## Replication Files
All data and code necessary to reproduce the reported results are licensed under the CC-BY-4.0 license and are publicly available as https://osf.io/yc3r4/?view_only=3ce3ca267c7047e0942940ebd08846d0. The analyses were performed in *R* [@rcore].

# Results
We detect numerous anomalous mortality and fertility events at the US state-level since 1999. A full listing of these anomalies can be found in the **Supplementary Materials**. We begin by highlighting significant fertility and mortality anomalies across all three types of outliers (Additive Outliers, Level Shift Outliers, and Temporary Change outliers) with plausible explanations. We then highlight two strong anomalies that belie explanation. Finally, we conclude with a summary of the anomalies we detect.

@@

conn_tstat <- prettyNum(outlier.fert.conn$outliers$tstat, digits = 3)
conn_births <- prettyNum(outlier.fert.conn$outliers$coefhat, digits = 0)
conn_tot <- prettyNum(abs(sum(outlier.fert.conn$y - outlier.fert.conn$yadj)), digits = 0, big.mark = ",")
conn_per <- percent(abs(sum(outlier.fert.conn$y - outlier.fert.conn$yadj) / sum(outlier.fert.conn$yadj[80:180])))

mag_la1 <- round(outlier.fert.la$outliers$coefhat[1],1)
mag_la2 <- round(outlier.fert.la$outliers$coefhat[2],1)
# int_la1 <- round(outlier.fert.la$outliers$coefhat[1] - confint(outlier.fert.la$fit, level=0.95)[5],1)
# int_la2 <- round(outlier.fert.la$outliers$coefhat[1] - confint(outlier.fert.la$fit, level=0.95)[6],1)
```

@@

```{r , include=FALSE}
n.sims <- 1e3

# This function will simulate 100 time series with similar values as the target database and make a ggplot figure version of tsoutliers baseplot but with an 95th percentile confidence interval.
makefigure <- function(this.outlier, titles){
set.seed(1)
sims.hist <- matrix(NA,nrow=n.sims,ncol=length(this.outlier$y))
# for ( ii in 1:n.sims ) {
#   # setWinProgressBar(pb,ii,paste(ii,"of",n.sims))
#   sims.hist[ii,] <- this.outlier$yadj+rnorm(length(this.outlier$yadj),0,sd(this.outlier$yadj))
#   model <- ets(ts(sims.hist[ii,],frequency=frequency(this.outlier$yadj)))
# }

   createsim <- function(ii){
    # set.seed(runif(1,1,1000))
    arima.sim(n=length(this.outlier$y), model=list(this.outlier$fit), sd = sdval) + as.vector(this.outlier$yadj)
   }
     sdval <- sqrt(this.outlier$fit$sigma2)
  for ( ii in 1:n.sims ) {
    sims.hist[ii,] <- as.vector(createsim())
  }

conf <- c(0.95)
ylim <- range(
  apply(sims.hist,2,quantile,prob=c((1-max(conf))/2,1-(1-max(conf))/2)))
boundaries.hist <- apply(sims.hist,2,quantile,prob=c((1-rev(conf)[1])/2,1-(1-rev(conf)[1])/2))

a <- data.frame(Y=as.matrix(this.outlier$yadj),
                date = time(this.outlier$yadj),
                ymin = as.matrix(boundaries.hist[1,]),
                ymax = as.matrix(boundaries.hist[2,])
)
b <- data.frame(y = as.matrix(this.outlier$y),
                date = time(this.outlier$y))
pts <- data.frame(date = this.outlier$times,
                  y = this.outlier$y[this.outlier$outliers$ind])

effectsizes <- data.frame(date =time(this.outlier$y),
                           y= this.outlier$y - this.outlier$yadj)

top <- ggplot(data= a, aes(x = date, y=Y)) +
  geom_line(data= b, aes(x=date, y=y), color = "red") +
  geom_line() +
  geom_ribbon(aes(ymin = ymin, ymax = ymax), alpha = 0.2) +
  geom_point(data=pts, aes(x=date, y =y), color = "red", fill = NA, size = 2, shape =21) +
  theme_bw() +
  scale_y_continuous(labels=comma) +
  labs(x="", y ="",
       title = paste(titles, "<br>
       <span style='font-size:10pt'><span style='color:#FF0000'>**Original**</span> and **Adjusted**  series</span>"
       # caption = "Shaded region is the 95th percentile confidence interval"
       )) +
  theme(
    plot.title = element_markdown(lineheight = 1.2),
    axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x = element_blank(),
    plot.margin = unit(c(0.1, 0, 0, 0), "cm")
  )

bot <- 
  ggplot(data = effectsizes, aes(x=date, y=y)) +
  geom_line(color="blue") +
  theme_bw() +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_y_continuous(labels=comma,
                     position = "right") +
  labs(x="", y ="")

together <- plot_grid(top, bot, ncol=1,
           align = "v", rel_heights = c(1,0.6))

return(together)
}

```{r FertilityAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.height = 6, fig.cap= paste("\\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.} The top part of each panel contains the original time series (light gray), the adjusted, counter-factual time series in the absence of anomalies (blue), and the red dots correspond to the onset of detected anomalies. The bottom part of each panel contains the magnitude and type of the outlier in red. In (a), we detect two outliers, back-to-back, likely resulting from Hurricane Katrina in August and October 2005, representing a decrease of more than 4,400 births due to the hurricane (t-statistics= -6.002 and -4.997). In (b), we detect one outlier, a level shift outlier (LS) in August 2009. We believe this reduction is attributable to the stock market crash 11 months earlier (t-statistic = -3.71). \\label{fig:fertla}")}

a<- makefigure(outlier.fert.la, "Louisiana - Fertility")
b <- makefigure(outlier.fert.conn, "Connecticut - Fertility")
# a <-  function() {
#   plot(outlier.fert.la, cex.axis=0.8)
#   title(main = "Louisiana - Fertility", adj = 0.9)
# }
# 
# b <- function(){
#   plot(outlier.fert.conn, cex.axis=0.8) +
#   title(main = "Connecticut - Fertility", adj = 0.88)
# }

# plot_grid(ggdraw(a) ,
#                ggdraw(b), 
#                ncol=1,
#                labels= "auto")

together <- plot_grid(a,
               b, 
               ncol=1,
               labels= "auto",
               rel_heights = c(1, 1))
ggdraw(add_sub(together, "Shaded region is the 95th percentile confidence interval", size = 8 ,x=1, hjust=1.1))
```
